Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs

### 2021-02-05 submission/manuscript.Rmd
Remove ROC curve analysis and replace with threshold analysis

```{r}
thresholds <- read_tsv(here("data/processed/thresholds_for_single_otu.tsv"),
											 col_types=cols(region = col_character(),
											 							 .default = col_double())) %>%
	filter(n_rrns == 7)

threshold_v19 <- thresholds %>% filter(region == "v19") %>% pull(threshold) * 100
threshold_v4 <- thresholds %>% filter(region == "v4") %>% pull(threshold) * 100
threshold_v34 <- thresholds %>% filter(region == "v34") %>% pull(threshold) * 100
threshold_v45 <- thresholds %>% filter(region == "v45") %>% pull(threshold) * 100

lump_split <- read_tsv(here("data/processed/lumped_split_rate.tsv"),
											 col_types = cols(region = col_character(),
											 								 .default = col_double()))

@@

lump3_v19 <- lump_split %>% filter(threshold ==0.03 & region == "v19") %>% pull(lump_rate) * 100
lump3_v4 <- lump_split %>% filter(threshold ==0.03 & region == "v4") %>% pull(lump_rate) * 100
lump3_v34 <- lump_split %>% filter(threshold ==0.03 & region == "v34") %>% pull(lump_rate) * 100
lump3_v45 <- lump_split %>% filter(threshold ==0.03 & region == "v45") %>% pull(lump_rate) * 100

```

A method to avoid splitting a single genome into multiple units of inference is to cluster 16S rRNA gene sequences together that are similar. However, this also increases the risk of lumping together genes from different species that are similar to each other. Therefore, I assessed the impact of the threshold used to define clusters of 16S rRNA genes on the propensity to split a genome apart or to lump species together. For full length 16S rRNA gene sequences, I found that at a threshold of `r format_pct(threshold_v19)`%, 95% of the species with 7 copies of the rrn operon would be represented by a single OTU. Similarly, thresholds of `r format_pct(threshold_v4)`, `r format_pct(threshold_v34)`, and `r format_pct(threshold_v45)`% were observed for the V4, V3-V4, and V4-V5 regions, respectively. However, at these thresholds, multiple species could be represented by the same OTU (**Figure 2**). At the highest level of resolution, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU when considering full length sequences and `r format_pct(lump3_v4)`, `r format_pct(lump3_v34)`, and `r format_pct(lump3_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. Given the risk of splitting a genome into multiple OTUs is more biologically problematic than lumping species together, larger thresholds are advisable.

```{r}
sensitivity_specificity <- read_tsv(here("data/processed/rrnDB.roc.tsv"),
													 col_types = cols(.default = col_double(),
													 								 region = col_character()
													 								 )
													 ) %>%
	mutate(sensitivity = true_pos / (true_pos + false_neg),
				 specificity = true_neg / (true_neg + false_pos)) %>%
	group_by(region, threshold) %>%
	summarize(sensitivity = median(sensitivity),
						specificity = median(specificity),
						.groups="drop") %>%
	select(region, threshold, sensitivity, specificity)

balance <- sensitivity_specificity %>%
	mutate(diff=abs(sensitivity - specificity)) %>%
	group_by(region) %>%
	summarize(min_diff = min(diff),
						threshold = threshold[which.min(diff)],
						sensitivity = sensitivity[which.min(diff)],
						specificity = specificity[which.min(diff)],
						.groups="drop"
						)

balance_v19 <- balance %>% filter(region == "v19") %>% pull(threshold) * 100
balance_v4 <- balance %>% filter(region == "v4") %>% pull(threshold) * 100
balance_v34 <- balance %>% filter(region == "v34") %>% pull(threshold) * 100
balance_v45 <- balance %>% filter(region == "v45") %>% pull(threshold) * 100

distance <- sensitivity_specificity %>%
	mutate(distance = sqrt((specificity - 1)^2 + (sensitivity - 1)^2)) %>%
	group_by(region) %>%
	summarize(min_distance = min(distance),
						threshold = threshold[which.min(distance)],
						sensitivity = sensitivity[which.min(distance)],
						specificity = specificity[which.min(distance)],
						.groups="drop"
						)

distance_v19 <- distance %>% filter(region == "v19") %>% pull(threshold) * 100
distance_v4 <- distance %>% filter(region == "v4") %>% pull(threshold) * 100
distance_v34 <- distance %>% filter(region == "v34") %>% pull(threshold) * 100
distance_v45 <- distance %>% filter(region == "v45") %>% pull(threshold) * 100

```

To provide a more nuanced approach to selecting a threshold, it would be useful to to quantify the sensitivity and specificity of characterizing bacterial species using OTUs defined at different thresholds. I created confusion matrices for multiple regions of the 16S rRNA gene: true positives were those cases where two ASVs were joined in the same OTU and the same species; true negatives were those cases where two ASVs from different OTUs came from different species; false positives were those ASVs that joined the same OTU, but were from different species; and false negatives were those ASVs that joined different OTUs, but were from the same species. By calculating the sensitivity and specificity for each threshold and each region of the 16S rRNA gene, I was able to constuct a receiver operator characteristic curve (ROC; **Figure S1**). Because the ROC curve represents a range of possible thresholds and sensitivities and specificities, I used two metrics to select the best threshold for defining an OTU. First, I identified the thresholds where the sensitivity and specificity were most similar to each other. For this criterion, the best distance thresholds were `r format_pct(balance_v19)`% (V1-V9), `r format_pct(balance_v4)`% (V4), `r format_pct(balance_v34)`% (V3-V4), and `r format_pct(balance_v45)`% (V4-V5). Second, I identified the distance threshold that resulted in the point on the ROC curve that was closest to perfect classification. For this criterion, the best distance thresholds were `r format_pct(distance_v19)`% (V1-V9), `r format_pct(distance_v4)`% (V4), `r format_pct(distance_v34)`% (V3-V4), and `r format_pct(distance_v45)`% (V4-V5). Surprisingly, these analyses revealed that thresholds near 3% distance balance the risks of splitting genomes into separate OTUs and lumping species into the same OTU.

@@

**Figure 1. The ratio of number of distinct ASVs per copy of the *rrn* operon increases for a species as the number of genomes sampled increases.** Each point represents a different species and is shaded to be 80% transparent so that when points overlap they become darker. The blue line represents a smoothed fit through the data.

@@

\includegraphics{../figures/roc_curve.pdf}

```{r}
asv_sens_spec <- sensitivity_specificity %>%
	filter(threshold == 0) %>% 
	mutate(sensitivity = format(sensitivity, digits=3, nsmall=3),
				 specificity = format(specificity, digits=3, nsmall=3),
				 pretty = paste(sensitivity, specificity, sep="/")
				 )
v19 <- asv_sens_spec %>% filter(region == "v19") %>% pull(pretty)
v34 <- asv_sens_spec %>% filter(region == "v34") %>% pull(pretty)
v4 <- asv_sens_spec %>% filter(region == "v4") %>% pull(pretty)
v45 <- asv_sens_spec %>% filter(region == "v45") %>% pull(pretty)

```

**Figure S1. Distance thresholds larger than 3% provide better sensitivity and specificity when assigning ASVs to OTUs when trying to represent species-level classifications.** The sensitivities and specificities for ASVs were `r v19`, `r v34`, `r v4`, and `r v45` for the V1-V9, V3-V4, V4, V4-V5 regions, respectively. The gray diagonal line represents the position of where the sensitivity and specificity were equal. These data represent the median fractions for sensitivity and specificity across 100 randomizations. In each randomization, one genome was sampled from each species.

>>>>

```{r}
thresholds <- read_tsv(here("data/processed/thresholds_for_single_otu.tsv"),
											 col_types=cols(region = col_character(),
											 							 .default = col_double())) %>%
								filter(n_genomes > 100)

threshold_v19 <- thresholds %>% filter(n_rrns == 7 & region == "v19") %>% pull(threshold) * 100
threshold_v4 <- thresholds %>% filter(n_rrns == 7 & region == "v4") %>% pull(threshold) * 100
threshold_v34 <- thresholds %>% filter(n_rrns == 7 & region == "v34") %>% pull(threshold) * 100
threshold_v45 <- thresholds %>% filter(n_rrns == 7 & region == "v45") %>% pull(threshold) * 100

three_v19 <- thresholds %>% filter(threshold <= 0.03 & region == "v19") %>% tail(n=1) %>% pull(n_rrns)
three_v4 <- thresholds %>% filter(threshold <= 0.03 & region == "v4") %>% tail(n=1) %>% pull(n_rrns)
three_v34 <- thresholds %>% filter(threshold <= 0.03 & region == "v34") %>% tail(n=1) %>% pull(n_rrns)
three_v45 <- thresholds %>% filter(threshold <= 0.03 & region == "v45") %>% tail(n=1) %>% pull(n_rrns)
```

A method to avoid splitting a single genome into multiple units of inference is to cluster together similar 16S rRNA gene sequences. Therefore, I assessed the impact of the distance threshold used to define clusters of 16S rRNA genes on the propensity to split a genome into separate clusters. I observed that as the number of copies of the *rrn* operon increased, the distance threshold required to reduce the ASVs in each genome to a single OTU increased (Figure 1). Among species with 7 copies of the *rrn* operon (e.g., *E. coli*), I found that a threshold of `r format_pct(threshold_v19)`% was required to reduce full length ASVs to a single OTU in 95% of the species. Similarly, thresholds of `r format_pct(threshold_v4)`, `r format_pct(threshold_v34)`, and `r format_pct(threshold_v45)`% were required for the V4, V3-V4, and V4-V5 regions, respectively. But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v4`, `r three_v34`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V4, V3-V4, and V4-V5 regions, respectively. Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters.

```{r}
lump_split <- read_tsv(here("data/processed/lumped_split_rate.tsv"),
											 col_types = cols(region = col_character(),
											 								 .default = col_double()))

@@

lump3_v19 <- lump_split %>% filter(threshold ==0.03 & region == "v19") %>% pull(lump_rate) * 100
lump3_v4 <- lump_split %>% filter(threshold ==0.03 & region == "v4") %>% pull(lump_rate) * 100
lump3_v34 <- lump_split %>% filter(threshold ==0.03 & region == "v34") %>% pull(lump_rate) * 100
lump3_v45 <- lump_split %>% filter(threshold ==0.03 & region == "v45") %>% pull(lump_rate) * 100
```

At broad thresholds thresholds multiple species could be represented by the same OTU (**Figure 2**). Using ASVs, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU when considering full length sequences and `r format_pct(lump3_v4)`, `r format_pct(lump3_v34)`, and `r format_pct(lump3_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. Considering that species designations are unevenly applied and reflect multiple biases, the risk of splitting a genome into multiple OTUs more problematic than clustering species together. Therefore, larger thresholds are advisable.

@@

\includegraphics{../figures/copy_number_threshold_plot.pdf}

**Figure 1. The distance threshold required to prevent the splitting of genomes into multiple OTUs increases as the number of *rrn* operons in the genome increases.** Each line represents the median distance threshold for each region of the 16S rRNA gene that is required for 95% of the species with the indicated numbrer of *rrn* operons to cluster their ASVs to a single OTU. The median distance threshold was calculated across 100 randomizations in which one genome was sampled from each species. Only those number of *rrn* operons that were found in more than 100 species are included.

@@

**Figure S1. The ratio of number of distinct ASVs per copy of the *rrn* operon increases for a species as the number of genomes sampled increases.** Each point represents a different species and is shaded to be 80% transparent so that when points overlap they become darker. The blue line represents a smoothed fit through the data.

### 2021-02-05 submission/manuscript.tex
Remove ROC curve analysis and replace with threshold analysis

A method to avoid splitting a single genome into multiple units of
inference is to cluster 16S rRNA gene sequences together that are
similar. However, this also increases the risk of lumping together genes
from different species that are similar to each other. Therefore, I
assessed the impact of the threshold used to define clusters of 16S rRNA
genes on the propensity to split a genome apart or to lump species
together. For full length 16S rRNA gene sequences, I found that at a
threshold of \textbf{5.5}\%, 95\% of the species with 7 copies of the
rrn operon would be represented by a single OTU. Similarly, thresholds
of \textbf{2.5}, \textbf{4.0}, and \textbf{3.5}\% were observed for the
V4, V3-V4, and V4-V5 regions, respectively. However, at these
thresholds, multiple species could be represented by the same OTU
(\textbf{Figure 2}). At the highest level of resolution, \textbf{3.6}\%
of the species shared a 16S rRNA gene sequence variant with another
species when considering full length sequences and \textbf{14.9},
\textbf{10.2}, and \textbf{12.0}\% when considering the V4, V3-V4, and
V4-V5 regions, respectively. At the commonly used 3\% threshold,
\textbf{25.2}\% of the species shared an OTU when considering full
length sequences and \textbf{33.0}, \textbf{29.4}, and \textbf{32.2}\%
when considering the V4, V3-V4, and V4-V5 regions, respectively. Given
the risk of splitting a genome into multiple OTUs is more biologically
problematic than lumping species together, larger thresholds are
advisable.

To provide a more nuanced approach to selecting a threshold, it would be
useful to to quantify the sensitivity and specificity of characterizing
bacterial species using OTUs defined at different thresholds. I created
confusion matrices for multiple regions of the 16S rRNA gene: true
positives were those cases where two ASVs were joined in the same OTU
and the same species; true negatives were those cases where two ASVs
from different OTUs came from different species; false positives were
those ASVs that joined the same OTU, but were from different species;
and false negatives were those ASVs that joined different OTUs, but were
from the same species. By calculating the sensitivity and specificity
for each threshold and each region of the 16S rRNA gene, I was able to
constuct a receiver operator characteristic curve (ROC; \textbf{Figure
S1}). Because the ROC curve represents a range of possible thresholds
and sensitivities and specificities, I used two metrics to select the
best threshold for defining an OTU. First, I identified the thresholds
where the sensitivity and specificity were most similar to each other.
For this criterion, the best distance thresholds were \textbf{6.0}\%
(V1-V9), \textbf{4.5}\% (V4), \textbf{5.5}\% (V3-V4), and \textbf{4.0}\%
(V4-V5). Second, I identified the distance threshold that resulted in
the point on the ROC curve that was closest to perfect classification.
For this criterion, the best distance thresholds were \textbf{5.5}\%
(V1-V9), \textbf{3.5}\% (V4), \textbf{4.5}\% (V3-V4), and \textbf{3.5}\%
(V4-V5). Surprisingly, these analyses revealed that thresholds near 3\%
distance balance the risks of splitting genomes into separate OTUs and
lumping species into the same OTU.

@@

\textbf{Figure 1. The ratio of number of distinct ASVs per copy of the
\emph{rrn} operon increases for a species as the number of genomes
sampled increases.} Each point represents a different species and is
shaded to be 80\% transparent so that when points overlap they become
darker. The blue line represents a smoothed fit through the data.

@@

\includegraphics{../figures/roc_curve.pdf}

\textbf{Figure S1. Distance thresholds larger than 3\% provide better
sensitivity and specificity when assigning ASVs to OTUs when trying to
represent species-level classifications.} The sensitivities and
specificities for ASVs were \textbf{0.435/1.000}, \textbf{0.748/1.000},
\textbf{0.887/0.999}, and \textbf{0.853/0.999} for the V1-V9, V3-V4, V4,
V4-V5 regions, respectively. The gray diagonal line represents the
position of where the sensitivity and specificity were equal. These data
represent the median fractions for sensitivity and specificity across
100 randomizations. In each randomization, one genome was sampled from
each species.

>>>>

A method to avoid splitting a single genome into multiple units of
inference is to cluster together similar 16S rRNA gene sequences.
Therefore, I assessed the impact of the distance threshold used to
define clusters of 16S rRNA genes on the propensity to split a genome
into separate clusters. I observed that as the number of copies of the
\emph{rrn} operon increased, the distance threshold required to reduce
the ASVs in each genome to a single OTU increased (Figure 1). Among
species with 7 copies of the \emph{rrn} operon (e.g., \emph{E. coli}), I
found that a threshold of \textbf{5.5}\% was required to reduce full
length ASVs to a single OTU in 95\% of the species. Similarly,
thresholds of \textbf{2.5}, \textbf{4.0}, and \textbf{3.5}\% were
required for the V4, V3-V4, and V4-V5 regions, respectively. But, if a
3\% distance threshold was used, then ASVs from genomes containing fewer
than \textbf{5}, \textbf{8}, \textbf{6}, and \textbf{6} copies of the
\emph{rrn} operon would reliably be clustered into a single OTU for ASVs
from the V1-V9, V4, V3-V4, and V4-V5 regions, respectively.
Consequently, these results demonstrate that broad thresholds must be
used to avoid splitting different operons from the same genome into
separate clusters.

At broad thresholds thresholds multiple species could be represented by
the same OTU (\textbf{Figure 2}). Using ASVs, \textbf{3.6}\% of the
species shared a 16S rRNA gene sequence variant with another species
when considering full length sequences and \textbf{14.9}, \textbf{10.2},
and \textbf{12.0}\% when considering the V4, V3-V4, and V4-V5 regions,
respectively. At the commonly used 3\% threshold, \textbf{25.2}\% of the
species shared an OTU when considering full length sequences and
\textbf{33.0}, \textbf{29.4}, and \textbf{32.2}\% when considering the
V4, V3-V4, and V4-V5 regions, respectively. Considering that species
designations are unevenly applied and reflect multiple biases, the risk
of splitting a genome into multiple OTUs more problematic than
clustering species together. Therefore, larger thresholds are advisable.

@@

\includegraphics{../figures/copy_number_threshold_plot.pdf}

\textbf{Figure 1. The distance threshold required to prevent the
splitting of genomes into multiple OTUs increases as the number of
\emph{rrn} operons in the genome increases.} Each line represents the
median distance threshold for each region of the 16S rRNA gene that is
required for 95\% of the species with the indicated numbrer of
\emph{rrn} operons to cluster their ASVs to a single OTU. The median
distance threshold was calculated across 100 randomizations in which one
genome was sampled from each species. Only those number of \emph{rrn}
operons that were found in more than 100 species are included.

@@

\textbf{Figure S1. The ratio of number of distinct ASVs per copy of the
\emph{rrn} operon increases for a species as the number of genomes
sampled increases.} Each point represents a different species and is
shaded to be 80\% transparent so that when points overlap they become
darker. The blue line represents a smoothed fit through the data.
